Prolfqua - Proteomics label free quantification using linear models.

### 2021-12-02 inst/MyArticle/ProlfquaPlain.Rmd
changes needed because prolfqua uses LazyLoad False.

---
journal: jacsat
type: article
author:
  - name: Witold E. Wolski
  - name: Christian Panse
  - name: Jonas Grossman
  - name: Paolo Nanni
  - name: Ralph Schlappbach
title:
  formatted: Prolfqua - \textsf{Pro}teomics label free quantification using linear models.
   class\footnote{A footnote for the title}
  short: prolfqua
abbr: IR,NMR,UV
keywords: American Chemical Society, \LaTeX
abstract: We introduce here the R package and benchmark linear modelling methods available in R for label free protein expression analysis.
bibliography: acstest.bib
csl: american-chemical-society.csl
output: rticles::acs_article
editor_options: 
  chunk_output_type: console
---

\begin{tocentry}
Some journals require a graphical entry for the Table of Contents.
This should be laid out ``print ready'' so that the sizing of the
text is correct.

Inside the \texttt{tocentry} environment, the font used is Helvetica
8\,pt, as required by \emph{Journal of the American Chemical
Society}.

The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
permitted for  \emph{Journal of the American Chemical Society}
graphical table of content entries. The box will not resize if the
content is too big: instead it will overflow the edge of the box.

This box and the associated title will always be printed on a
separate page at the end of the document.
\end{tocentry}

@@

The prolfqua package fits the linear model to all the proteins in the dataset. Afterward, we can compute contrast to test the hypothesis, perform ANOVA, or model selection using the likelihood ratio test for thousands of proteins. By exploiting the parallel structure of the analysis, _prolfqua_ implements t-statistic moderation to improve inference for small sample sizes [_limma_]. It also computes probabilities of differential protein regulation based on peptide level models [_ropeca_].

@@

## Benchmarking
The benchmark dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentration [@shen2018ionstar]. We know that for _H. sapiens_ proteins the difference $\beta$ between two dilutions should be $\beta = 0$ while for _E. coli_ proteins we know that the difference between dilutions should be $\beta \ne 0$.

@@

For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)).
From the confusion matrix we can compute measures such as true positive rate (TPR), false positive rate (FPR), false discovery proportion (FDP) or accuracy (ACC) which are given by:

```{r confusionMatrix}
table <- data.frame( c("beta != 0", "beta == 0", "Total"), matrix(c("TP","FP","R","FN","TN","","P","N","m"), ncol = 3, byrow = T))
colnames(table) <- c("Prediction \\ Truth","E.coli", "H.sapiens", "Total")
knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, P - all positive cases (all E. coli proteins), N - all negative cases (all H. sapiens proteins), m- all proteins.", format="latex")

@@

For some of the proteins, quantification results are missing in some conditions. All we required when attempting to fit a model for a protein is that there are at least two peptides per protein quantified in any of the twenty samples. That means that for some proteins, not all model parameters are estimable. 
In case of the linear model, the contrast function we implemented can handle partial models and provides estimates of some of the contrasts (see Table \@ref(tab:tablesMedpolish)). For the mixed effect models, the used `lmerTest::contest` function does not support models with missing parameter estimates, and hence we have 4 (all) or 0 (no) contrast estimates for a protein (see Table \@ref(tab:tablesMM)). For the Bayesian models obtained with _brms_ or _rstan_ we also only computed contrasts for models with complete fixed effect estimates (see Tables \@ref(tab:countsBrms), \@ref(tab:countsBrmsStudent) and \@ref(tab:countsRstand)). The ROPECA approach starts by fitting a model to each peptide and the contrast function can handle missing parameter estimates (see Table \@ref(tab:tablesRopeca)).
These differences result in different numbers of fold-change estimates for each type of model, i.e. in a different set of ground truths, which makes comparing model performances involved.
To deal with these differences, we computed our benchmark estimates, including the peptides and proteins, without fold-change estimates, by accordingly updating the numbers in the confusion matrix. To compute the $FPR$ and $TPR$ including the missing estimates we set $P =$ #all _E. coli_ peptides/proteins detected, and $N =$ #all _H. sapiens_ proteins detected.

@@

$$
z' = z \times 1/N\sum_{i=1}^N S_i
$$
We decided to apply this simple transformation because it only requires to estimate two parameters per sample and works for experiments with thousands of proteins as well as for experiments where only a few hundreds of proteins per sample are measured.

@@

tt
```

@@

tt <- tt %>% dplyr::filter(contrastsName  %in% relevantContrasts)
tt <- select(tt, all_of(c("contrasts","ratio", "expected fold change")))
knitr::kable(tt, caption = "Contrasts used for benchmark.", format = "latex")

@@

## Linear model

Here we show how the two-stage modelling by first estimating the protein intensities using the Tukey median polish and then fitting a linear model performs.

@@

@@

* (`prot_med_lm`) : linear model fitted to protein estimates obtained using Tukey's median polish 
* (`prot_lme4`) : mixed effects model fitted using _lme4_ 
* (`prot_ROPECA`) : linear models fitted on peptide level and aggregation of moderated p-values using the beta distribution
* (`prot_rstan`) : mixed effects model fitted with _rstan_ and with prior over $\varepsilon$ 
* (`prot_brms`) : mixed effects model fitted with _brms_ and default priors 
* (`prot_brms_student_t`) : mixed effects model fitted with _brms_ and shrinkage prior over fixed effects coefficients .

A relevant parameter is the number of proteins for which we estimated the contrasts (see Table \@ref(tab:completeCasesTab)). It indicates how robust the models are in the presence of missing data. We observe, that when using the linear model on protein level intensities or the bayesian version of the mixed effects models we were able to estimate for 4024 (out of the 4099) proteins all the four examined contrasts (complete cases). When using the mixed effect model implementation in _lme4_ this number drops to $4013$, while when summarizing peptide level models it drops further to $3999$ proteins.

The different number of proteins for which we obtained statistics of contrasts makes comparing the scores for various models complicated because the set of proteins differs, e.g. $4024$ versus $3999$. 
Therefore, we incorporate the failure to generate statistics for contrasts into the benchmark by setting $P$ and $N$ equal to the number of all proteins (see Table \@ref(tab:confusionMatrix)) and then computing the $FPR$ and $TPR$.

## Outline

# Extra information when writing JACS Communications

@@

The authors thank technology platform fund of the University of Zurich.

@@

\end{suppinfo}
## References

>>>>

---
title: "Prolfqua - Proteomics label free quantification using linear models."
author: 'Witold Wolski Std Nr:'
date: "6/11/2021"
output: 
  bookdown::html_document
bibliography: bibliography.bib

---

# Abstract

@@

The prolfqua package fits the linear model to all the proteins in the dataset. Afterward, we can compute contrast to test the hypothesis, perform ANOVA, or model selection using the likelihood ratio test for thousands of proteins. By exploiting the parallel structure of the analysis, _prolfqua_ implements t-statistic moderation to improve inference for small sample sizes [@Smyth2004linear]. It also computes probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced].

@@

## Benchmarking

The benchmark dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentration [@shen2018ionstar]. We know that for _H. sapiens_ proteins the difference $\beta$ between two dilutions should be $\beta = 0$ while for _E. coli_ proteins we know that the difference between dilutions should be $\beta \ne 0$.

@@

For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)).
From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), false discovery proportion (FDP) or accuracy (ACC) which are given by:

```{r confusionMatrix}
table <- data.frame( c("beta != 0", "beta == 0", "Total"), matrix(c("TP","FP","R","FN","TN","","P","N","m"), ncol = 3, byrow = T))
colnames(table) <- c("Prediction \\ Truth","E.coli", "H.sapiens", "Total")
knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, P - all positive cases (all E. coli proteins), N - all negative cases (all H. sapiens proteins), m- all proteins.")

@@

For some of the proteins, quantification results are missing in some conditions. All we required when attempting to fit a model for a protein is that there are at least two peptides per protein quantified in any of the twenty samples. That means that for some proteins, not all model parameters are estimable. 
In case of the linear model, the contrast function we implemented can handle partial models and provides estimates of some of the contrasts. For the mixed effect models, the used `lmerTest::contest` function does not support models with missing parameter estimates, and hence we have either four or zero contrast estimates for a protein. The ROPECA approach starts by fitting a model to each peptide and the contrast function can handle missing parameter estimates.
These differences result in different numbers of fold-change estimates for each type of model, i.e. in a different set of ground truths, which makes comparing model performances involved.

@@

$$
z' = z \times 1/N\sum_{i=1}^N S_i
$$
We decided to apply this simple transformation because it only requires to estimate two parameters per sample. For the ionstar dataset we estimated \bar{x} and $S$ using the human proteins only.

@@

@@

tt <- tt %>% dplyr::filter(contrastsName  %in% relevantContrasts)
tt <- dplyr::select(tt, all_of(c("contrasts","ratio", "expected fold change")))
knitr::kable(tt, caption = "Contrasts used for benchmark.")

@@

## Linear models

We fitted a linear model implemented by the R function `lm` to protein intensities inferred from peptide intensities using the Tukey's median polish. The second mixed-effects model, implemented in the R function `lmer`, fitted to peptide level intensities. We modeled the peptide measurements as repeated measurements of the protein. The third model is again a linear model but this time fitted to peptide intensities. By this we have for each protein several models, which we then summarize.

## Contrast estimation with imputation

To handle cases, where no observations were made in one of the conditions, we model the condition estimates and variances as follows. Either we use the mean of the protein intensity in the group, or if there are no observations we impute by using the average of $10\%$ smallest group mean intensities in the experiment. Here we assume that if there are no observations it is because of the limit of detection (LOD). To estimate the variance, we compute the pooled variance and standard deviation of all conditions $\hat{var}$. We assume that the variance of the protein is constant in all the conditions. The t-statistic is then given by $\frac{\bar x}{\hat{\sigma}}$.

## Computing contrasts

Given a linear model contrasts can be computed by $\hat{\beta_{c}} = \sum l\beta_{m}$ and $\textrm{var} \hat\beta_c = l\sigma^2 (X^T X)^{-1} l^T$, with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of coefficients. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model.
For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `lmerTest` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom.

## P-value moderation

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom.  [@Smyth2004linear] discuss how, using the empirical Bayes paradigm, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior degrees of freedom, and posterior $\tilde \sigma$. These can be used to moderate the t-statistics by:
$\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}$
and p-values.

## Summarizing peptide level models

To summarize peptide level models to protein models we did applied the method suggested by [@Suomi2017bEnhanced], that is to use the median scaled p-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

To obtain the median p-value of the protein, we first rescaled the peptide p-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

\begin{equation}
p_{s} =
  \begin{cases}
1-p, & \textrm{if}~ \hat{\beta} > 0\\
p-1, & \textrm{otherwise}
\end{cases}
\end{equation}

Afterwards, the median scaled p-value $\tilde{p}_s$ is determined and using the transformation below, transformed back onto the original scale.

$$
\tilde{p} = 1 - |\tilde{p}_{s}|
$$
Because we used the median as the i-th order statistic  $i = n/2 +0.5$. Therefore, $\gamma = i = n/2 + 0.5$ and $\delta = n - i + 1 = n - (n/2 + 0.5) + 1 = n/2 + 0.5 = \gamma$ are used to parameterize the CDF of the Beta distribution.

@@

## Method comparison

@@

* (`prot_med_lm`) : linear model fitted to protein estimates obtained using Tukey's median polish 
* (`prot_lme4`) : mixed effects model fitted using _lme4_ to peptide intensities.
* (`prot_ROPECA`) : linear models fitted on peptide level and aggregation of moderated p-values using the beta distribution

The table below summarizes the contrast estimates produced which will be benchmarked.

|                    | Model                   | Contrast   | Moderation   | Aggregation  |
|--------------------|:-----------------------:|------------|--------------|--------------|
| Protein Intensity  |  lm                     | o          |  o           |              |
| Protein Intensity Imputed | pooled variance  | o          |  o           |              |
| Peptide Intensity  |  lmer                   | o          |  o           |              |
| Peptide Intensity  |  lm                     |            |              |  o           |

A relevant parameter is the number of proteins for which we estimated the contrasts (see Table \@ref(tab:completeCasesTab)). It indicates how robust the models are in the presence of missing data. We observe, that when using the linear model on protein level intensities or the bayesian version of the mixed effects models we were able to estimate for 4024 (out of the 4099) proteins all the four examined contrasts (complete cases). When using the mixed effect model implementation in _lme4_ this number drops to $4013$, while when summarizing peptide level models it drops further to $3999$ proteins. The different number of proteins for which we obtained statistics of contrasts makes comparing the scores for various models complicated because the set of proteins differs, e.g. $4024$ versus $3999$. 
Therefore, we incorporate the failure to generate statistics for contrasts into the benchmark by setting $P$ and $N$ equal to the number of all proteins (see Table \@ref(tab:confusionMatrix)) and then computing the $FPR$ and $TPR$.

```{r}
allBenchmarks <- readRDS("../../inst/Benchresults/allBenchmarks.RDS")
benchmark_msstats <- readRDS("../../inst/Benchresults/benchmark_msstats.RDS")
msFragger <- readRDS(file = "../../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

allBenchmarks$benchmark_mssstats <- benchmark_msstats
allBenchmarks$benchmark_msFragger <- msFragger

names(allBenchmarks)
allBenchmarks <- allBenchmarks[c("benchmark_imputation","benchmark_ProtModerated",  "benchmark_mixedModerated", "benchmark_ropeca","benchmark_merged","benchmark_mssstats", "benchmark_msFragger"   )]
```

```{r  fig.cap="Number and percentage of estimated contrasts by modelling method.", fig.width=5, fig.height=7}

dd <- map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd %>% mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd %>% group_by(name) %>% summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

nrgg <- dds %>% ggplot(aes(x = name, y = nrcontrasts )) + 
  geom_bar(stat = "identity", fill="white", colour = "black") + 
  coord_cartesian(ylim = c(min(dds$nrcontrasts) - 100, max(dds$nrcontrasts) + 10)) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5)) +
  geom_text(aes(label = round(nrcontrasts, digits = 1)),
            vjust = 1, hjust = -0.2, angle = -90) + 
  geom_text(aes(label = paste0("(",round(percent, digits = 1),"%)")),
            vjust = -1, hjust = -0.2, angle = -90) 
nrgg

```{r fig.cap="Partial area under the ROC curve at 10% FPR. Red line average area under the curve.", fig.width=10, fig.width=10}

ttt <- sapply(allBenchmarks, function(x){x$complete(FALSE)})
res <- map_df(allBenchmarks, function(x){x$pAUC()})
res <- res %>% mutate(whatfix = case_when(what == "scaled.beta.based.significance" ~ "scaled.p.value", TRUE ~ what))

res$contrast %>% unique()
#resAllB <- res %>% dplyr::filter(contrast == "dilution_(9/7.5)_1.2")
#resAllB <- res %>% dplyr::filter(contrast == "dilution_(4.5/3)_1.5")
#resAllB <- res %>% dplyr::filter(contrast == "dilution_(7.5/6)_1.25")

norm <- res %>% group_by(contrast,whatfix) %>% summarize(meanpAUC_10 = mean(pAUC_10))
res <- inner_join(res, norm)
res <- mutate(res , pAUC_10n = pAUC_10 - meanpAUC_10)

resAllB <- res %>% dplyr::filter(contrast == "all")

p1 <- ggplot(resAllB, aes(x = Name, y = pAUC_10)) +
  geom_bar(stat = "identity") +
  facet_wrap(~whatfix)  + 
  coord_cartesian(ylim = c(min(resAllB$pAUC_10),max(resAllB$pAUC_10))) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5)) +
  geom_hline(aes(yintercept = meanpAUC_10), color="red")
p1
```

```{r fig.cap="Difference to mean partial area under the ROC curve for various models, at 10% FPR, by fold change.", fig.width=10, fig.width=10}

p2 <- ggplot(res, aes(x = contrast, y = pAUC_10n, group = Name)) +
  geom_line(stat = "identity",aes(linetype = Name, color = Name)) + 
  facet_wrap(~whatfix, scales = "free") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5)) +
  geom_hline(aes(yintercept = 0), color = "red")

p2
```

```{r FDRfdp, fig.cap = "Compare FDR estimate with false discovery proportion (FDP).", fig.width=8, fig.height=8}
dd <- map_df(allBenchmarks, function(x){res <- x$get_confusion_FDRvsFDP(); res$name <- x$model_name;res})
ddb <- filter(dd, contrast == "dilution_(4.5/3)_1.5")
ddb <- dd %>% dplyr::filter(contrast == "dilution_(7.5/6)_1.25")
ddb <- dd %>% dplyr::filter(contrast == "all")

ddb %>% ggplot(aes(y = FDP_,  x  = scorecol )) + 
  geom_line(aes(color = model_name, linetype = model_name)) +
  facet_wrap(~contrast) + 
   geom_abline(intercept = 0, slope = 1, color = 2)

```{r fig.cap="Distribution of the FDR values produced."}
dd %>% ggplot(aes(x = scorecol)) + geom_histogram() + facet_wrap(~model_name)
```

@@

The authors thank the technology platform fund of the University of Zurich.

@@

\end{suppinfo}

## References
